---
title: "Team 10"
format: html
editor: visual
---



## Clean Code for 

```{r}
# Load libraries
library(dplyr)
library(tidyr)
```

```{r}
suicide_data <- read.csv("C:/Users/hanzi/Downloads/Death_rates_for_suicide__by_sex__race__Hispanic_origin__and_age__United_States.csv")
drug_overdose_data <- read.csv("C:/Users/hanzi/Downloads/Drug_overdose_death_rates__by_drug_type__sex__age__race__and_Hispanic_origin__United_States.csv")
```

```{r}
colnames(suicide_data)
colnames(drug_overdose_data)
```


```{r}
head(suicide_data)
head(drug_overdose_data)
```
```{r}
# View structure
glimpse(suicide_data)
glimpse(drug_overdose_data)

# Display first few rows
head(suicide_data)
head(drug_overdose_data)
```

```{r}
unique(suicide_data$STUB_LABEL)
unique(suicide_data$STUB_NAME)
unique(suicide_data$AGE)
unique(drug_overdose_data$STUB_LABEL)
unique(drug_overdose_data$STUB_NAME)
unique(drug_overdose_data$AGE)
```

```{r}
# Clean the suicide dataset
suicide_clean <- suicide_data %>%
  mutate(
    Gender = case_when(
      grepl("Male", STUB_LABEL) ~ "Male",
      grepl("Female", STUB_LABEL) ~ "Female",
      TRUE ~ "All persons"
    ),
    Race = case_when(
      grepl("White", STUB_LABEL) ~ "White",
      grepl("Black or African American", STUB_LABEL) ~ "Black or African American",
      grepl("Asian or Pacific Islander", STUB_LABEL) ~ "Asian or Pacific Islander",
      grepl("American Indian or Alaska Native", STUB_LABEL) ~ "American Indian or Alaska Native",
      TRUE ~ "Unknown"
    )
  ) %>%
  select(
    Gender, 
    Race,
    Year = YEAR,
    Age_Group = AGE,
    Suicide_Rate = ESTIMATE
  ) %>%
  filter(Gender != "All persons") # Exclude general rows if needed

# Clean the drug overdose dataset
drug_overdose_clean <- drug_overdose_data %>%
  mutate(
    Gender = case_when(
      grepl("Male", STUB_LABEL) ~ "Male",
      grepl("Female", STUB_LABEL) ~ "Female",
      TRUE ~ "All persons"
    ),
    Race = case_when(
      grepl("White", STUB_LABEL) ~ "White",
      grepl("Black or African American", STUB_LABEL) ~ "Black or African American",
      grepl("Asian or Pacific Islander", STUB_LABEL) ~ "Asian or Pacific Islander",
      grepl("American Indian or Alaska Native", STUB_LABEL) ~ "American Indian or Alaska Native",
      TRUE ~ "Unknown"
    )
  ) %>%
  select(
    Gender,
    Race,
    Year = YEAR,
    Age_Group = AGE,
    Overdose_Rate = ESTIMATE
  ) %>%
  filter(Gender != "All persons") # Exclude general rows if needed
```


```{r}
# Check for missing data
sum(is.na(suicide_clean))
sum(is.na(drug_overdose_clean))

# Remove rows with missing data
suicide_clean <- suicide_clean %>% drop_na()
drug_overdose_clean <- drug_overdose_clean %>% drop_na()
```
## Data Provenance Description

The data for this research was sourced from Data.gov, the U.S. government's open data platform that provides access to a wide range of datasets collected by federal agencies. Specifically, the datasets on drug overdose death rates and suicide death rates were compiled by the National Center for Health Statistics (NCHS), a division of the Centers for Disease Control and Prevention (CDC). These datasets were created to track trends in mortality rates across different demographic groups and to inform public health strategies. The purpose of using these datasets is to address the research question: Which gender-race combinations show the highest vulnerability to either suicide or drug overdose as a cause of death? The data includes death rates per 100,000 residents, categorized by gender, race, age group, and year, offering a detailed demographic breakdown. This information provides the foundation for analyzing disparities and identifying groups that may be disproportionately affected.

## FAIR Principles

The datasets used in this research align with the FAIR Principles—Findability, Accessibility, Interoperability, and Reusability. These datasets are highly findable, as they are hosted on Data.gov, a widely recognized and centralized government data repository, with detailed metadata that enables easy discovery through search engines. They are accessible to the public without cost or restriction, provided in a standard CSV format that can be downloaded and analyzed using common tools. The datasets are interoperable, as they follow widely accepted standards for data organization, including clear labels, consistent structures, and metadata, which make them easy to integrate with other datasets or software. Finally, they are reusable, as they come with clear attribution to the National Center for Health Statistics (NCHS) and are accompanied by descriptions of their purpose, collection methodology, and scope, ensuring that researchers can repurpose the data for further studies or new questions while maintaining its integrity.

## Attributes for Analysis

The analysis focuses on the following attributes: Gender, Race, Suicide_Rate, and Overdose_Rate. These attributes are directly available in the datasets and provide the foundation for examining patterns and disparities. Gender and Race provide demographic context, while Suicide_Rate and Overdose_Rate represent the number of deaths per 100,000 residents for suicides and drug overdoses, respectively. These attributes will be used to identify the demographic combinations (e.g., "White males") most affected by either suicide or drug overdose, answering the research question with clear and actionable insights.


## Research Topic: Exploring Suicide and Drug Overdose Rates Across Different Population Groups

This study aims to analyze disparities in suicide and drug overdose rates, focusing on demographic factors such as race, gender, and age.

#### Research Question 1: Which gender-race combinations show the highest vulnerability to either suicide or drug overdose as a cause of death?

Suicide and drug overdoses often have intersecting risk factors, such as socioeconomic status and mental health. Examining gender and race intersectionally provides a understanding of vulnerabilities.

#### Research Question 2: Is there a correlation between suicide rates and drug overdose rates in different racial demographic groups?

This question investigates whether suicide rates and drug overdose rates are statistically related within different racial demographic groups. The aim is to explore whether the prevalence of these two public health crises is interconnected, potentially driven by shared underlying causes or social determinants.

```{r}
# Merge the two datasets
merged_data <- merge(
  suicide_clean,
  drug_overdose_clean,
  by = c("Gender", "Race", "Year", "Age_Group"),
  all = FALSE 
)

merged_data <- merged_data %>%
  filter(Race != "Unknown") %>% # Remove rows with unknown race
  select(-Age_Group, -Year)     # Drop the Age_Group and Year columns entirely

```

```{r}
write.csv(merged_data, "Cleaned_Combined_Death_Rates.csv", row.names = FALSE)
```

```{r}
summary(merged_data)

# Check unique values
table(merged_data$Gender)
table(merged_data$Race)
table(merged_data$Year)
table(merged_data$Age_Group)
```

```{r}
write.csv(
  merged_data,
  "C:/Users/hanzi/Downloads/Cleaned_Combined_Death_Rates.csv",
  row.names = FALSE
)

print("The cleaned and further filtered dataset has been saved successfully!")
```

## Exploratory Data Analysis (EDA)

```{r}
library(ggplot2)
# Bar chart: Death rates by gender and race
ggplot(data = merged_data, aes(x = Race, y = Suicide_Rate, fill = Gender)) +
  geom_bar(stat = "identity", position = "dodge") +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1, size = 10),
    axis.title.x = element_text(size = 12, face = "bold"),
    axis.title.y = element_text(size = 12, face = "bold")
  ) +
  labs(
    title = "Suicide Rates by Race and Gender",
    x = "Race",
    y = "Suicide Rate",
    fill = "Gender"
  )
```



```{r}
# Scatter plot: Comparing suicide and overdose rates
ggplot(merged_data, aes(x = Suicide_Rate, y = Overdose_Rate, color = Gender)) +
  geom_point() +
  labs(title = "Comparison of Suicide and Overdose Rates",
       x = "Suicide Rate",
       y = "Overdose Rate",
       color = "Gender") +
  theme_minimal()
```

```{r}
# Load required library
library(tidyr)

# Transform data to long format
merged_data_long <- merged_data %>%
  pivot_longer(cols = c(Suicide_Rate, Overdose_Rate), 
               names_to = "Variable", 
               values_to = "Rate")

# Heatmap: Combined rates by race and gender
library(ggplot2)

ggplot(data = merged_data_long, aes(x = Race, y = Gender, fill = Rate)) +
  geom_tile(color = "white") +
  facet_wrap(~ Variable, scales = "free", ncol = 2) + # Adjust facets
  scale_fill_gradient(low = "lightblue", high = "darkblue", name = "Rate") +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1, size = 10), # Rotate and space x-axis labels
    axis.title.x = element_text(size = 12, face = "bold"),
    axis.title.y = element_text(size = 12, face = "bold"),
    strip.text = element_text(size = 12, face = "bold") # For facet titles
  ) +
  labs(
    title = "Heatmap of Suicide and Overdose Rates by Race and Gender",
    x = "Race",
    y = "Gender"
  )
```

```{r}
# Summarize the data by Race and Gender
summary_table <- merged_data %>%
  group_by(Race, Gender) %>%
  summarise(
    Avg_Suicide_Rate = mean(Suicide_Rate, na.rm = TRUE),
    Avg_Overdose_Rate = mean(Overdose_Rate, na.rm = TRUE)
  )

# Display the summary table
library(knitr)
kable(summary_table, 
      col.names = c("Race", "Gender", "Average Suicide Rate", "Average Overdose Rate"),
      caption = "Summary Table of Average Suicide and Overdose Rates by Race and Gender")
```

```{r}
# QQ Plot for Suicide Rate
qqnorm(merged_data$Suicide_Rate, main = "QQ Plot of Suicide Rate")
qqline(merged_data$Suicide_Rate, col = "red", lwd = 2)
library(ggplot2)

# QQ Plot for Suicide Rate
ggplot(merged_data, aes(sample = Suicide_Rate)) +
  stat_qq() +
  stat_qq_line(color = "red") +
  labs(
    title = "QQ Plot of Suicide Rate",
    x = "Theoretical Quantiles",
    y = "Sample Quantiles"
  ) +
  theme_minimal()
```

```{r}
#| echo: false
#Correlation

# Load necessary libraries
library(dplyr)

# Assuming your dataset is `merged_data` with columns `Suicide_Rate`, `Overdose_Rate`, and `Race`

# Group by Race and calculate correlation for each racial group
correlation_results <- merged_data %>%
  group_by(Race) %>%
  summarize(
    correlation = cor(Suicide_Rate, Overdose_Rate, use = "complete.obs"),
    p_value = cor.test(Suicide_Rate, Overdose_Rate, use = "complete.obs")$p.value
  )

# Print the correlation results
print(correlation_results)


```




```{r codeAppend, ref.label=knitr::all_labels(), echo=TRUE, eval=FALSE}
      
```